---
title: "Spatial Inequalities in Educational Infrastructure: Bayesian Hierarchical Modelling of School Availability Across North Borneo"
# "Spatial Inequalities in Educational Infrastructure: Bayesian Hierarchical Modelling of School Availability Across North Borneo"
author:
  - name: Alvin Bong
    email: alvinbongjl@gmail.com
    # affiliations:
    #   - name: Universiti Brunei Darussalam
    #     id: ubd
    #     department: Mathematical Sciences, Faculty of Science
    #     address: Universiti Brunei Darussalam, Jalan Tungku Link
    #     city: Bandar Seri Begawan
    #     country: Brunei
    #     postal-code: BE 1410
    attributes:
      corresponding: true
date-modified: last-modified
bibliography: 
  - refs.bib
abstract: |
  Understanding the spatial distribution of educational infrastructure is essential for ensuring equitable access to schooling. This study examines regional disparities in school availability across Brunei and the neighboring Malaysian states of Sarawak and Sabah, with a focus on identifying potential underserved areas. First, exploratory data analysis (EDA) was conducted to compare school counts and student-teacher ratios at the district level across the three regions. The core analysis employs Standardized Incidence Ratio (SIR) and Bayesian spatial Poisson model using Integrated Nested Laplace Approximation (INLA) to estimate the relative abundance of schools across Brunei’s districts, adjusting for expected counts based on population, region size and socioeconomic indicator (house price + partially simulated). Spatially structured and unstructured random effects were incorporated to account for latent spatial processes. Posterior estimates identified districts with significantly lower school availability than the national baseline, supporting future policy planning and school placement. 
---

```{r}
#| label: setup
#| include: false
library(tidyverse)

library(tidyverse)
library(ggplot2)
library(mapview)
# Reorder OpenStreetMap as default
all_basemaps <- c("OpenStreetMap", 
                  "CartoDB.Positron",  
                  "CartoDB.DarkMatter", 
                  "Esri.WorldImagery", 
                  "Esri.WorldStreetMap") 
# Set default basemap to OSM
mapviewOptions(basemaps = all_basemaps)
mapviewOptions(fgb = FALSE)
library(bruneimap)
library(geodata)
library(sf)
library(viridis)
library(spdep)
library(INLA)
library(leaflet)
library(leaflet.extras2)
library(RColorBrewer)

# Data: map ---------------------------------------------------------------------
# Download Malaysia boundaries from GADM level1=States level=2Districts (Daerah)
mys_state <- geodata::gadm(country = "MYS", level = 1, path = tempdir())
mys_dis <- geodata::gadm(country = "MYS", level = 2, path = tempdir())
# Filter only North Borneo
mys_state_sf <-
  st_as_sf(mys_state) %>%
  select(state = NAME_1) %>%
  filter(state %in% c("Sarawak", "Sabah"))
sbh_state_sf <- mys_state_sf %>% filter(state == "Sabah")
swk_state_sf <- mys_state_sf %>% filter(state == "Sarawak")
mys_dis_sf <- 
  st_as_sf(mys_dis) %>% 
  select(state = NAME_1, district = NAME_2)
sbh_dis_sf <- mys_dis_sf %>% filter(state == "Sabah")
swk_dis_sf <- mys_dis_sf %>% filter(state == "Sarawak")
east_mys_dis_sf <- mys_dis_sf %>% filter(state %in% c("Sarawak", "Sabah"))

# Data: Primary & secondary Schools, Population, sf ----------------------------------------------------------------
# A. Malaysia
mys_sch_df <- read_csv("source/mys_schools_district.csv")

sbh_sch_df <- mys_sch_df %>% 
  filter(state == "Sabah", 
         date == "2018-01-01",
         stage %in% c("primary", "secondary"),
         district != "All Districts") %>% 
  group_by(district) %>% 
  summarise(schools = sum(schools, na.rm = TRUE))

# Telupid was formerly part of the Beluran District and is now a separate 
# administrative district with Telupid Town as its capital. 
# Fix: merge to old map
sbh_sch_df <- sbh_sch_df %>%
  mutate(district = case_when(
    district == "Telupid" ~ "Beluran",
    TRUE ~ district
  )) %>% 
  group_by(district) %>% 
  summarise(schools = sum(schools, na.rm = TRUE))
setdiff(sbh_dis_sf$district, sbh_sch_df$district)
setdiff(sbh_sch_df$district, sbh_dis_sf$district)

swk_sch_df <-
  mys_sch_df %>% 
  filter(state == "Sarawak", 
         date == "2018-01-01",
         stage %in% c("primary", "secondary"),
         district != "All Districts") %>% 
  group_by(district) %>% 
  summarise(schools = sum(schools, na.rm = TRUE))

swk_sch_df <- swk_sch_df %>%
  mutate(district = case_when(
    district == "Maradong" ~ "Meradong",
    district == "Kabong" ~ "Saratok",
    district == "Pusa" ~ "Saratok",
    district == "Sebauh" ~ "Bintulu",
    district == "Subis" ~ "Miri",
    TRUE ~ district
  )) %>% 
  group_by(district) %>% 
  summarise(schools = sum(schools, na.rm = TRUE))
setdiff(swk_dis_sf$district, swk_sch_df$district)
setdiff(swk_sch_df$district, swk_dis_sf$district)

# Join Population
mys_census2021 <- read_csv("source/mys_population_district.csv")
mys_census2021 <- mys_census2021 %>% 
  as_tibble() %>% 
  filter(ethnicity=="overall",
         date=="2021-01-01",
         age=="overall",
         sex=="both",
         state %in% c("Sarawak", "Sabah")) %>% 
  select(state, district, population) %>% 
  mutate(population = population*1000)
swk_census2021 <- mys_census2021 %>% 
  filter(state=="Sarawak") %>%
  select(district, population) %>% 
  mutate(district = case_when(
    district == "Maradong" ~ "Meradong",
    district == "Kabong" ~ "Saratok",
    district == "Pusa" ~ "Saratok",
    district == "Sebauh" ~ "Bintulu",
    district == "Subis" ~ "Miri",
    district == "Beluru" ~ "Miri",
    district == "Bukit Mabong" ~ "Kapit",
    district == "Matu" ~ "Mukah",
    district == "Pakan" ~ "Sarikei",
    district == "Selangau" ~ "Sibu",
    district == "Tanjung Manis" ~ "Mukah",
    district == "Tebedu" ~ "Serian",
    district == "Telang Usan" ~ "Miri",
    TRUE ~ district
  )) %>% 
  group_by(district) %>% 
  summarise(population = sum(population, na.rm = TRUE))
sbh_census2021 <- mys_census2021 %>% 
  filter(state=="Sabah") %>% 
  select(district, population) %>% 
  mutate(district = case_when(
    district == "Telupid" ~ "Beluran",
    district == "Kalabakan" ~ "Tawau",
    TRUE ~ district
  )) %>% 
  group_by(district) %>% 
  summarise(population = sum(population, na.rm = TRUE))

setdiff(sbh_sch_df$district, sbh_census2021$district)
setdiff(sbh_census2021$district, sbh_sch_df$district)
setdiff(swk_sch_df$district, swk_census2021$district)
setdiff(swk_census2021$district, swk_sch_df$district)

# Join sf
sbh_sch_df <- left_join(sbh_sch_df, sbh_census2021, by="district")
swk_sch_df <- left_join(swk_sch_df, swk_census2021, by="district")
sbh_sch_sf <- left_join(sbh_dis_sf, sbh_sch_df, by="district") %>% select(-state)
swk_sch_sf <- left_join(swk_dis_sf, swk_sch_df, by="district") %>% select(-state)
east_mys_sch_sf <- rbind(sbh_sch_sf, swk_sch_sf)

# mapview(east_mys_sch_sf, zcol="schools")



# B. Brunei
# Fix school on water village (slightly out of bound)
brn_sch_sf <- bruneimap::sch_sf %>% 
  mutate(district = case_when(
    School == "Sekolah Rendah Tanjong Kindana" ~ "Brunei-Muara",
    TRUE ~ district
  )) %>% 
  mutate(mukim = case_when(
    School == "Sekolah Rendah Tanjong Kindana" ~ "Mukim Kota Batu",
    TRUE ~ mukim
  ))

# Filter Only Primary, Secondary Government Schools
brn_mkm_sch_df <- brn_sch_sf %>% 
  filter(Sector == "MOE") %>% 
  select(School, Education.Level, kampong, mukim, district) %>% 
  filter(!Education.Level %in% c("Vocational / Technical Education", 
                                 "Higher Education",
                                 "Pre-primary",
                                 "Technical / Vocational Institution",
                                 "Vocational / Technical  Education")) %>% 
  group_by(mukim) %>% 
  summarise(schools = n()) %>% 
  st_drop_geometry()

brn_dis_sch_df <- brn_sch_sf %>% 
  filter(Sector == "MOE") %>%  # government schools
  select(School, Education.Level, kampong, mukim, district) %>% 
  filter(!Education.Level %in% c("Vocational / Technical Education", 
                                 "Higher Education",
                                 "Pre-primary",
                                 "Technical / Vocational Institution",
                                 "Vocational / Technical  Education")) %>% 
  group_by(district) %>% 
  summarise(schools = n()) %>% 
  st_drop_geometry()

# Join population
mkm_pop <- bruneimap::census2021 %>% 
  group_by(mukim) %>% 
  summarise(population = sum(population, na.rm = TRUE))
dis_pop <- bruneimap::census2021 %>% 
  mutate(district = case_when(
    district == "Brunei Muara" ~ "Brunei-Muara",
    TRUE ~ district
  )) %>% 
  group_by(district) %>% 
  summarise(population = sum(population, na.rm = TRUE))

# Join sf
brn_mkm_sch_sf <- left_join(mkm_sf, brn_mkm_sch_df, by="mukim") 
brn_dis_sch_sf <- left_join(dis_sf, brn_dis_sch_df, by="district") %>% select(district, schools)
brn_mkm_sch_sf <- left_join(brn_mkm_sch_sf, mkm_pop, by="mukim")
brn_dis_sch_sf <- left_join(brn_dis_sch_sf, dis_pop, by="district")

# C. North Borneo
nborneo_sch_sf <- rbind(east_mys_sch_sf, brn_dis_sch_sf)

# EDA1: school count ------------------------------------------------------------------
nborneo_sch_sf <- nborneo_sch_sf %>% mutate(area = as.numeric(st_area(geometry)))
nborneo_sch_sf <- nborneo_sch_sf %>% mutate(sch_pop = schools/population*1000,
                                            sch_area = schools/area * 1000000)

pal <- colorRampPalette(brewer.pal(9, "YlOrRd"))
# sch
m1 <- mapview(nborneo_sch_sf, zcol="schools", col.regions = pal, layer.name="School Count")
# sch:pop per 1000
m2 <- mapview(nborneo_sch_sf, zcol="sch_pop", col.regions = pal, layer.name="School per 1000 people")
#sch:area
m3 <- mapview(nborneo_sch_sf, zcol="sch_area", col.regions = pal, layer.name="School per km^2")
leafsync::sync(m1, m2, m3)

# EDA2: std_tcr --------------------------------------------------------
# student teacher ratio
brn_tchr <- bruneimap::tchr %>% 
  mutate(teachers = as.numeric(M) + as.numeric(`F`),
         district = District,
         stage = case_when(
           `Education Level` == "Primary" ~ "primary",
           `Education Level` == "Secondary" ~ "secondary",
           TRUE ~ `Education Level`)) %>% 
  filter(Sector == "MOE",
         stage %in% c("primary", "secondary")) %>% 
  select(district, stage, teachers)

brn_enrolment <- bruneimap::enrolment %>% 
  mutate(students = as.numeric(M) + as.numeric(`F`),
         district = District,
         stage = case_when(
           `Education Level` == "Primary" ~ "primary",
           `Education Level` == "Secondary" ~ "secondary",
           TRUE ~ `Education Level`)) %>% 
  filter(Sector == "MOE",
         stage %in% c("primary", "secondary")) %>% 
  select(district, stage, students)

mys_tchr <- read_csv("source/mys_teacher.csv")
mys_enrolment <- read_csv("source/mys_enrolment.csv")
east_mys_tchr <- mys_tchr %>% 
  filter(state %in% c("Sarawak", "Sabah"),
         stage %in% c("primary", "secondary"),
         district != "All Districts",
         sex == "both",
         date == "2018-01-01") %>% 
  select(-sex, -date, -state) %>% 
  mutate(district = case_when(
    district == "Telupid" ~ "Beluran",
    district == "Kalabakan" ~ "Tawau",
    district == "Maradong" ~ "Meradong",
    district == "Kabong" ~ "Saratok",
    district == "Pusa" ~ "Saratok",
    district == "Sebauh" ~ "Bintulu",
    district == "Subis" ~ "Miri",
    district == "Beluru" ~ "Miri",
    district == "Bukit Mabong" ~ "Kapit",
    # district == "Matu" ~ "Mukah",
    # district == "Pakan" ~ "Sarikei",
    # district == "Selangau" ~ "Sibu",
    district == "Tanjung Manis" ~ "Mukah",
    district == "Tebedu" ~ "Serian",
    district == "Telang Usan" ~ "Miri",
    TRUE ~ district
  )) %>% 
  group_by(district, stage) %>% 
  summarise(teachers = sum(teachers, na.rm = TRUE))
east_mys_enrolment <- mys_enrolment %>% 
  filter(state %in% c("Sarawak", "Sabah"),
         stage %in% c("primary", "secondary"),
         district != "All Districts",
         sex == "both",
         date == "2018-01-01") %>% 
  select(-sex, -date, -state) %>% 
  mutate(district = case_when(
    district == "Telupid" ~ "Beluran",
    district == "Kalabakan" ~ "Tawau",
    district == "Maradong" ~ "Meradong",
    district == "Kabong" ~ "Saratok",
    district == "Pusa" ~ "Saratok",
    district == "Sebauh" ~ "Bintulu",
    district == "Subis" ~ "Miri",
    district == "Beluru" ~ "Miri",
    district == "Bukit Mabong" ~ "Kapit",
    # district == "Matu" ~ "Mukah",
    # district == "Pakan" ~ "Sarikei",
    # district == "Selangau" ~ "Sibu",
    district == "Tanjung Manis" ~ "Mukah",
    district == "Tebedu" ~ "Serian",
    district == "Telang Usan" ~ "Miri",
    TRUE ~ district
  )) %>% 
  group_by(district, stage) %>% 
  summarise(students = sum(students, na.rm = TRUE))

setdiff(east_mys_tchr$district, east_mys_dis_sf$district)
setdiff(east_mys_dis_sf$district, east_mys_tchr$district)
setdiff(east_mys_enrolment$district, east_mys_dis_sf$district)
setdiff(east_mys_dis_sf$district, east_mys_enrolment$district)
# mys_enrolment[mys_enrolment$district %in% c("Matu", "Pakan", "Selangau"),]
# some changes in administartive boudnaries in 2017/2018

brn_std_tchr <- left_join(brn_enrolment, brn_tchr, by=c("district"="district", "stage"="stage"))
brn_std_tchr_sf <- brn_std_tchr %>% 
  mutate(district = case_when(
    district == "Brunei Muara" ~ "Brunei-Muara",
    TRUE ~ district
  )) %>% 
  left_join(., dis_sf, by="district") %>% 
  select(-id, -X, -Y, -perimeter, -area)

east_mys_std_tchr <- left_join(east_mys_enrolment, east_mys_tchr, by=c("district"="district", "stage"="stage"))
east_mys_std_tchr_sf <- east_mys_std_tchr %>% 
  left_join(., east_mys_dis_sf, by="district") %>% 
  select(-state)

nborneo_std_tchr_sf <- rbind(brn_std_tchr_sf, east_mys_std_tchr_sf)
nborneo_std_tchr_sf <- nborneo_std_tchr_sf %>% 
  mutate(std_tcr=students/teachers)

std_tcr_primary <- nborneo_std_tchr_sf %>%  filter(stage == "primary") %>%  st_as_sf()
std_tcr_secondary <- nborneo_std_tchr_sf %>%  filter(stage == "secondary") %>%  st_as_sf()

at <- seq(min(std_tcr_primary$std_tcr), max(std_tcr_primary$std_tcr), length.out = 5)
mapview(std_tcr_primary, zcol="std_tcr", col.regions = pal, at = at,
        color = "#3b3b3b",
        layer.name="Primary - std:tcr",
        highlight = leaflet::highlightOptions(weight = 4),
        popup = leafpop::popupTable(dplyr::mutate_if(std_tcr_primary, is.numeric,
                                                     round, digits = 2),
                                    zcol = c("district", "students", "teachers", "std_tcr"),
                                    row.numbers = FALSE, feature.id = FALSE))

at <- seq(min(std_tcr_secondary$std_tcr, na.rm = TRUE), 
          max(std_tcr_secondary$std_tcr, na.rm = TRUE), length.out = 5)
mapview(std_tcr_secondary, zcol="std_tcr", col.regions = pal, at = at,
        color = "#3b3b3b",
        layer.name="Secondary - std:tcr",
        highlight = leaflet::highlightOptions(weight = 4),
        popup = leafpop::popupTable(dplyr::mutate_if(std_tcr_secondary, is.numeric,
                                                     round, digits = 2),
                                    zcol = c("district", "students", "teachers", "std_tcr"),
                                    row.numbers = FALSE, feature.id = FALSE))
# Putatan NA <= only 1 primary school, 0 secondary



# 1. Global Spatial Autocorrelation ---------------------------------------


# 2. Model y ~ pop + hp (socioeconomic) + u_i +v_i (Only Brunei, by mukim) --------------------------

# hp
hp <- read_csv("source/brn_house_price.csv")
hp <- hp %>% 
  group_by(mukim) %>%
  summarise(price = median(price, na.rm = TRUE))
setdiff(mkm_sf$mukim, hp$mukim)

# replace NA schools <= 0
brn_mkm_sch_sf$schools[is.na(brn_mkm_sch_sf$schools)] <- 0

brn_mkm_sch_sf <- left_join(brn_mkm_sch_sf, hp, by = "mukim")
brn_mkm_sch_sf <- brn_mkm_sch_sf %>% 
  mutate(hp = price) %>% 
  select(-X, -Y, -perimeter, -area, -price, -id)





# Fix missing hp using INLA (gaussian)
# Other options: 
# educated guess based on location & population
test <- brn_mkm_sch_sf #refer to appendix
#   Alt 1. drop missing data (would lost info)
#   Alt 2. Fill in using mean of neighbours (need loop, complicated, some neighbours also NA)
nb <- poly2nb(brn_mkm_sch_sf)
nb2INLA("map.adj", nb)
g <- inla.read.graph(filename = "map.adj")
brn_mkm_sch_sf$re_u <- 1:nrow(brn_mkm_sch_sf)
brn_mkm_sch_sf$log.hp <- log(brn_mkm_sch_sf$hp)
formula <- log.hp ~ f(re_u, model = "bym2", graph = g)
res <- inla(formula, family="gaussian", data=brn_mkm_sch_sf, 
            control.predictor=list(compute=TRUE),
            control.compute = list(return.marginals.predictor = TRUE))

res$summary.fitted.values$mean # values too high for forested area
brn_mkm_sch_sf$PM <- res$summary.fitted.values[, "mean"]

# Transformation marginals with inla.tmarginal()
marginals <- lapply(res$marginals.fitted.values,
                    FUN = function(marg){inla.tmarginal(function(x) exp(x), marg)})

# Obtain summaries of the marginals with inla.zmarginal()
marginals_summaries <- lapply(marginals,
                              FUN = function(marg){inla.zmarginal(marg)})

# Posterior mean
brn_mkm_sch_sf$PMoriginal <- sapply(marginals_summaries, '[[', "mean") 
# Replace missing hp with predicted
brn_mkm_sch_sf$hp[is.na(brn_mkm_sch_sf$hp)] <- brn_mkm_sch_sf$PMoriginal[is.na(brn_mkm_sch_sf$hp)]






# SIR (bad, overhighlihts, since low school count)
# brn_mkm_sch_sf <- brn_mkm_sch_sf %>%  filter(schools>3) # optional, to avoid extremes
brn_mkm_sch_sf$area <- as.numeric(st_area(brn_mkm_sch_sf))
brn_mkm_sch_sf$Y <- brn_mkm_sch_sf$schools
brn_mkm_sch_sf$E <- sum(brn_mkm_sch_sf$schools)/sum(brn_mkm_sch_sf$population) * brn_mkm_sch_sf$population
brn_mkm_sch_sf$SIR <- brn_mkm_sch_sf$Y/brn_mkm_sch_sf$E
at <- c(0,0.5,1,2,3,4,5)
mapview(brn_mkm_sch_sf, zcol="SIR", col.region=pal, at=at, layer.name="SIR") # bad, overhighlihts, since low school count




# fit model y ~ pop + area + hp (socioeconomic) + u_i +v_i
# Concern: Is school count too little. If poisson, ok.
nb <- poly2nb(brn_mkm_sch_sf)
nb2INLA("map.adj", nb)
g <- inla.read.graph(filename = "map.adj")
brn_mkm_sch_sf$re_u <- 1:nrow(brn_mkm_sch_sf)
brn_mkm_sch_sf <- brn_mkm_sch_sf %>%
  mutate(
    pop_s = population / 1000,   # per 1000 people
    area_s = as.numeric(area) / 1000000,  # e.g., km^2 instead of m^2
    hp_s = hp / 1000              # per 1000 currency units
  )

formula <- Y ~ pop_s + area_s + hp_s + f(re_u, model = "bym2", graph = g)
# formula <- Y ~ pop_s + hp_s + f(re_u, model = "bym2", graph = g)
# formula <- Y ~ pop_s + f(re_u, model = "bym2", graph = g)
# formula <- Y ~ hp_s + f(re_u, model = "bym2", graph = g)

res <- inla(formula, family = "poisson", data = brn_mkm_sch_sf, E=E,
            control.predictor = list(compute = TRUE),
            control.compute = list(return.marginals.predictor = TRUE),
            verbose = TRUE)

res$summary.fixed

brn_mkm_sch_sf$RA <- res$summary.fitted.values[, "mean"]

m1 <- mapview(brn_mkm_sch_sf, zcol = "RA", col.region=pal, at=at)
m2 <- mapview(brn_mkm_sch_sf, zcol = "hp", col.region=pal)
at <- c(0,100,1000,10000,20000)
m3 <- mapview(brn_mkm_sch_sf, zcol = "population", col.region=pal, at=at)
leafsync::sync(m1, m2, m3)

summary(res)

# Use Exceedance Prob.
brn_mkm_sch_sf$exc <- sapply(res$marginals.fitted.values,
                    FUN = function(marg){inla.pmarginal(q = 0.75, marginal = marg)})

at <- c(0,0.25,0.5,0.75,1)
mapview(brn_mkm_sch_sf, zcol = "exc", col.region=pal, at=at, 
        layer.name="Exceedance Probability RA < 0.8")
```

## Introduction

Education is a foundational pillar of human and national development, influencing social well-being, economic growth, and long-term sustainability. Ensuring equitable access to education through sufficient infrastructure, fair resource distribution, and balanced student–teacher ratios is critical to delivering quality learning experiences.

The global significance of education is recognized in Sustainable Development Goal 4, which promotes inclusive and equitable quality education for all [1]. Nationally, Brunei Darussalam’s national vision, Wawasan Brunei 2035, positions education as a cornerstone of the country's long-term development goals. 

While several studies have examined general aspects of education in Brunei, there have been limited studies based on quantitative spatial methods, with only one examining the spatial distribution and hotspots of schools. This project aims to addresses that gap by first conducting a comparative analysis of school availability and student–teacher ratios across Brunei’s districts, with additional context from neighboring Malaysian states, Sarawak and Sabah. Given that Brunei has higher PISA scores than Malaysia, it would be interesting to explore whether accessibility may be a potential factor. / Malaysia has more population than Brunei, how does the school compare

Next, Standardized Incidence Ratios (SIR) and Bayesian hierarchical models are used to identify districts where school availability falls significantly below the national baseline,  supporting future policy planning future school placements.

<!-- The rest of this paper is organized as follows. In section -->
<!-- [2](#sec-method){reference-type="ref" reference="sec-method"}, we -->
<!-- describe the methodology used. In section -->
<!-- [3](#sec-results){reference-type="ref" reference="sec-results"}, the -->
<!-- results are shown. In section [4](#sec-conc){reference-type="ref" -->
<!-- reference="sec-conc"}, we conclude. -->

## Data
In this study, we only focus on primary and secondary public schoools. School dataset 2018 is used becaused Although newer school statistics exist, they are only published in summary form. Thus, we use the 2018 school dataset, which provides more granular spatial detail necessary for district- and mukim-level aggregation.Population census data from 2021 was used despite the mismatch in year with the school data, as Brunei conducts a national census only once every ten years, and 2021 is the most recent available.

Various data variables sourced were joined using left_join() and merged using rbind() after preprocessing, sourced:

### Brunei
Data on school locations, student–teacher ratios, administrative boundaries, and population were sourced from the `bruneimap` R package. The school dataset (sch_sf) includes georeferenced point data for each institution. For this analysis, we filtered the dataset to include only government primary and secondary schools, to align with available school data in Malaysia.

Despite composing of only four districts, district-level aggregation was used for broader comparisons, such as student–teacher ratios and school counts, to match the smallest available administrative unit in Malaysian data. Mukims ($N = 39$), a finer administrative level, were used for the Bayesian spatial analysis of school availability. 

To incorporate a socioeconomic indicator, median house of each region data from approximately 30,000 property listings spanning 1993–2025. The median house price was calculated per mukim to represent local socioeconomic variation. Some mukims had missing data; these values were imputed using predictions from an INLA-based Gaussian model. Manual corrections based on local knowledge were initially tested, but INLA-predicted values were ultimately used, as both methods yielded similar model outcomes. Given the nature of the dataset, house prices were treated as partially simulated estimates.

### Malaysia
Malaysian data were sourced from the national open data portal [data.gov.my], including district-level school counts and population estimates for Sarawak and Sabah. The school data were filtered to include government primary and secondary schools only, for comparability with Brunei data.

Administrative boundary (districts) were obtained via the `geomdata` R package, as OpenStreetMap (`osmdata`) does not provide administrative divisions at the required level of granularity.

Some inconsistencies were found between school and boundary data, especially where older districts had been split into newer ones. In these cases, school counts were only available for the original (larger) districts. To maintain consistency, we excluded the newer subdivisions and manually recoded schools in affected areas to align them with the closest valid district.



## Method {#sec-method}

Exploratory data anlysis using Clorepath maps for schools count, by area (usingst_area) studnet teacher ratio

Neighbour type

POisson INLA hierarchical model

covariates: 
house price as socio economic index

Spatial Poisson model
Justify offset (population)
Explain fixed and random effects
Introduce INLA + BYM


## Results {#sec-results}

- School count quite similar, more in main city
- studnet teacher ratio, primary simiilar other than dense city (miri, kuching, kota kinabalu), but brunei has lesser secondary std-tcr
- There are significantly fewer schools in Mukim xxx

```{r}

pal <- colorRampPalette(brewer.pal(9, "YlOrRd"))
# sch
m1 <- mapview(nborneo_sch_sf, zcol="schools", col.regions = pal, layer.name="School Count")
# sch:pop per 1000
m2 <- mapview(nborneo_sch_sf, zcol="sch_pop", col.regions = pal, layer.name="School per 1000 people")
#sch:area
m3 <- mapview(nborneo_sch_sf, zcol="sch_area", col.regions = pal, layer.name="School per km^2")
leafsync::sync(m1, m2, m3)

at <- seq(min(std_tcr_primary$std_tcr), max(std_tcr_primary$std_tcr), length.out = 5)
mapview(std_tcr_primary, zcol="std_tcr", col.regions = pal, at = at,
        color = "#3b3b3b",
        layer.name="Primary - std:tcr",
        highlight = leaflet::highlightOptions(weight = 4),
        popup = leafpop::popupTable(dplyr::mutate_if(std_tcr_primary, is.numeric,
                                                     round, digits = 2),
                                    zcol = c("district", "students", "teachers", "std_tcr"),
                                    row.numbers = FALSE, feature.id = FALSE))

at <- seq(min(std_tcr_secondary$std_tcr, na.rm = TRUE), 
          max(std_tcr_secondary$std_tcr, na.rm = TRUE), length.out = 5)
mapview(std_tcr_secondary, zcol="std_tcr", col.regions = pal, at = at,
        color = "#3b3b3b",
        layer.name="Secondary - std:tcr",
        highlight = leaflet::highlightOptions(weight = 4),
        popup = leafpop::popupTable(dplyr::mutate_if(std_tcr_secondary, is.numeric,
                                                     round, digits = 2),
                                    zcol = c("district", "students", "teachers", "std_tcr"),
                                    row.numbers = FALSE, feature.id = FALSE))
# Putatan NA <= only 1 primary school, 0 secondary

at <- c(0,0.5,1,2,3,4,5)
mapview(brn_mkm_sch_sf, zcol="SIR", col.region=pal, at=at, layer.name="SIR") # bad, overhighlihts, since low school count

m1 <- mapview(brn_mkm_sch_sf, zcol = "RA", col.region=pal, at=at)
m2 <- mapview(brn_mkm_sch_sf, zcol = "hp", col.region=pal)
at <- c(0,100,1000,10000,20000)
m3 <- mapview(brn_mkm_sch_sf, zcol = "population", col.region=pal, at=at)
leafsync::sync(m1, m2, m3)

at <- c(0,0.25,0.5,0.75,1)
mapview(brn_mkm_sch_sf, zcol = "exc", col.region=pal, at=at, 
        layer.name="Exceedance Probability RA < 0.8")
```

## Discussion & Limitation

There are significantly fewer schools in Mukim xxx, when explored, its not forested area. Housing region, some may be newer nieghbourhood so.. good idea for mroe schools there, nearer from home, less transportation cost, environmental, time. 

Limitation:
only public schools, age groups, 2018 data, school types, 
some simulation in housing price,
house price is listing price (only a proxy for market values), may deviate from actual sales price due to negotian dynamics and other factors
## Conclusions {#sec-conc}

- School count quite similar, more in main city
- studnet teacher ratio, primary simiilar other than dense city (miri, kuching, kota kinabalu), but brunei has lesser secondary std-tcr
- There are significantly fewer schools in Mukim xxx

## References {.unnumbered}

::: {#refs}
:::